---
title: "fastPLS: Mathematical Methods, Implementations, and User Guide"
output: BiocStyle::html_document
header-includes:
    - \usepackage{array}
vignette: >
    %\VignetteIndexEntry{fastPLS Mathematical Methods and User Guide}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(width = 68)
library(fastPLS)
```

## Overview

`fastPLS` combines partial least squares (PLS) modelling with truncated singular
value decomposition (SVD), compiled numerical implementations, and optional
accelerator execution. This vignette first introduces the mathematical methods,
then explains their CPU, CUDA, and Metal implementations, and finally presents
the public analysis workflow.

## Mathematical Methods and Implementations

### Singular Value Decomposition Algorithms

`fastPLS` separates the PLS model from the singular-vector algorithm used
inside the fit. fastPLS uses its native randomized SVD (rSVD) implementation
following the randomized range-finder strategy of Halko, Martinsson, and Tropp
(2011). In the compiled CPU implementation, a Gaussian test matrix is generated,
the target matrix is sketched, optional power iterations are applied, the sketch
is orthonormalized by an economic QR decomposition, and the final singular
vectors are obtained from a much smaller projected matrix. When the randomized
sketch dimension reaches the smaller matrix dimension, or when one
matrix dimension is very small, the code falls back to an exact SVD because a
truncated randomized approximation would no longer reduce the computation.

Native rSVD is the only public SVD solver. It is approximate: finite factors
alone do not establish accurate singular directions or predictive agreement.
Standalone `fastsvd()` reports singular-triplet checks and effective controls.
If ordinary CPU sketches fail their checks, native rSVD retries with wider
fresh sketches and additional power iterations. Failed or over-budget recovery
raises an error. The recovery budget concerns major numerical buffers, not
complete-process memory.

Sequential PLS fitting reports its seed, effective controls, execution profile,
finite-factor checks and direction rule. These diagnostics do not establish
equivalence to an independent PLS implementation. Compare several seeds and,
when feasible, an independent reference before interpreting sensitive
coefficients or latent subspaces. Increasing `power` or `oversample` can improve
an approximation but is not a guarantee of agreement. Native rSVD is the
only solver exposed by this package.
PLS-SVD and standalone `fastsvd()` use `oversample = 32`, `power = 5` by
default. Accelerated SIMPLS, OPLS, and kernel-PLS use 32 oversampling
directions and five power iterations for ordinary matrix shapes. Numeric
responses with at least 64 columns and a response-to-sample ratio of at least
0.2 use 48 directions and six iterations. Classification with at least 32
classes and no more than 20 samples per class uses 64 directions and seven
iterations. When the explicit predictor-response cross-covariance would exceed
512 MiB, the massive-matrix profile requests `oversample = 12` and `power = 1`;
executed controls and refresh width are recorded separately. CPU, Metal, and
ordinary CUDA SIMPLS-family routes generate a seeded sketch from the current
deflated cross-covariance. On a massive problem, CPU, Metal, and CUDA float64
use a rank-one randomized direction at each refresh. CUDA float32 can obtain up
to eight independently initialized candidates per refresh through component 64
before returning to rank-one refreshes; CUDA keeps the sequential state on the
device. For eligible large dummy-coded classification, CPU, CUDA, and Metal can
instead obtain a candidate block of up to 64 directions. Candidates are
accepted sequentially through the SIMPLS orthogonalization and deflation
equations. Because the cross-covariance changes after every accepted component,
the block route is an approximate sequential estimator and is not claimed to
equal a fresh dominant-direction solve at every step. The explicit and
matrix-free routes evaluate the same cross-covariance operator.

For large predictor-response cross-products, `fastPLS` can avoid forming
`S = X^T Y`. This matrix-free route is called `xprod` in the code. The SVD
backend evaluates products with `S` and `S^T` only when needed, for example
`Sv = X^T(Yv)` and `S^Tu = Y^T(Xu)`. This preserves the same mathematical
operator while reducing memory pressure. For classification, labels can be used
as PLS-DA responses without explicitly materializing a large dense one-hot
matrix in the label-aware paths.

### Partial Least Squares Algorithms

`method = "plssvd"` implements the direct PLS-SVD formulation, in which the
dominant singular subspace of the predictor-response cross-covariance matrix is
computed once and reused for all requested component counts. This makes PLS-SVD
efficient when the response rank is the limiting dimension. The implementation
adds compact latent prediction factors: instead of always storing and applying
a full coefficient array, prediction can be performed through low-rank factors
based on the latent projection and response-side coefficients. The cached
score-Gram route forms the score cross-product once and reuses its leading
blocks for the requested prefixes. If the centred cross-covariance is
approximated by $S \approx UDV^\top$, scores are $T_a=XU_a$ and the
response map solves $(T_a^\top T_a)H_a=D_aV_a^\top$. Predictions are
$X_{\mathrm{new}}U_aH_a$, with the training response mean restored.
Singular directions alone are therefore not regression coefficients.
PLS-SVD can also
use the matrix-free `xprod` route for randomized SVD when the
cross-covariance matrix would be large.

`method = "simpls"` is the optimized SIMPLS implementation and is the default
general-purpose PLS core. The statistical structure follows de Jong's SIMPLS
algorithm (de Jong, 1993): each component is obtained from the current
cross-covariance state, converted into a score and loading direction, and
followed by SIMPLS orthogonalization and deflation. The rSVD
route generates candidates from a newly randomized sketch at each refresh of
the current deflated operator; a refresh can supply one or several candidates.
Like `pls::simpls.fit`, one fit provides the sequential path through all
components up to `ncomp`; fastPLS does not claim this standard path construction
as a novelty. Cross-covariance deflation is updated after each accepted component.
The component-wise CPU rSVD route derives a reproducible component-specific seed
and generates a new Gaussian sketch at each refresh. Metal initializes one
reproducible random sequence per fit and draws a new sketch for each component.
CUDA generates newly initialized randomized candidates in resident device
workspaces. Eligible large dummy-coded classification fits can generate up to
64 directions per CPU, CUDA, or Metal decomposition and consume
them sequentially.
The CPU route also evaluates candidate scores and predictor loadings in matrix
blocks. Eligibility depends on predictor, response and sample dimensions, and
the requested component count; it is not determined by sample count alone.
This batched approximate route need not reproduce a fresh dominant-direction
solve after each deflation. The ordinary CPU regression route remains
component-wise; CUDA float32 additionally has a bounded massive-response block
route. The C++ core
also caches the rank-one
deflation product, updates coefficient and fitted-value paths incrementally,
and uses compact prediction factors. For low-dimensional high-sample problems,
the code
can cache cross-products such as `X^T X` and `X^T Y`; for large response or
predictor dimensions, it can instead use matrix-free `xprod` products. These
routes are selected by matrix shape: they reduce avoidable work or storage in
their intended regimes but are not asserted to make every SIMPLS problem
faster.

When the response dimension is no larger than the predictor dimension and is
at most 512, the randomized direction calculation can use the smaller
response-side Gram matrix $S^\top S$. After each accepted component, fastPLS
updates this matrix with the corresponding rank-one deflation product. This
avoids repeatedly sketching the full tall cross-covariance while leaving the
subsequent SIMPLS score, loading, orthogonalization, and deflation steps
unchanged. The CPU implementation uses BLAS/LAPACK operations, CUDA keeps the
Gram matrix and its updates in device memory for float32 and float64, and Metal
uses its native float32 matrix operators.

`method = "opls"` implements orthogonal PLS following the orthogonal signal
correction idea of Trygg and Wold (2002). It estimates response-orthogonal
variation in `X`, subtracts this structured variation, stores the orthogonal
filter, and then fits the optimized PLS core to the filtered matrix. The same
filter is applied to new samples at prediction time.

`method = "kernelpls"` follows kernel PLS as described by Rosipal and Trejo
(2001). Linear kernel PLS is dispatched to the ordinary linear PLS core to avoid
unnecessary kernel materialization. For radial-basis and polynomial kernels,
the training kernel is centered and the selected PLS core is fitted in kernel
space. The same user-facing `pls()` function therefore covers linear PLS-SVD,
SIMPLS, OPLS, and kernel PLS while sharing the accelerated low-rank machinery.

### Classification Methods

Classification is handled as PLS-DA followed by one of two decoding heads.
`classifier = "argmax"` uses the largest predicted dummy-response score and is
the standard PLS-DA decoding rule. `classifier = "lda"` fits a regularized
linear discriminant analysis model in the PLS score space following Fisher's
discriminant principle (Fisher, 1936).

### Computational Backends

The `backend` argument selects where the main numerical work is performed.
`backend = "cpu"` uses the package's compiled, dependency-free C++ core with
native rSVD. It is usually best for small and moderate
datasets, low-dimensional problems, and machines without GPU support.
The CPU kernels use Apple Accelerate by default on macOS. Linux and Windows
prefer OpenBLAS when configuration finds it through `OPENBLAS_ROOT`,
`pkg-config`, or Rtools; otherwise the package uses the BLAS/LAPACK supplied by
R. Set `FASTPLS_USE_OPENBLAS=1` before installation to require OpenBLAS.
`fastPLS_blas()` reports the library selected when the installed package was
compiled. Linux and Windows performance benchmarks should proceed only when
this function returns `"OpenBLAS"`.
Linking against multithreaded OpenBLAS enables eligible
matrix operations to use multiple CPU cores. For example,
`options(backend = "cpu", n.cores = 4L)` selects CPU execution and requests four
threads from the linked BLAS/OpenMP runtime. Explicit `backend` and `n.cores`
arguments override their respective session options. The same resolved
`n.cores` value is retained by nested folds, refits, permutations, and
predictions. The sequential PLS deflation steps remain serial,
so additional cores do not guarantee a speed improvement for every matrix
shape.
Additional BLAS threads help only when the matrix products are sufficiently
large; sequential updates and thread-management overhead can limit or reverse
the benefit. Compare timings on the intended workload rather than assuming
that more threads are always faster.

For CPU self-products, fastPLS selects GEMM or the symmetric rank-k operation
SYRK from the linked numerical library according to platform, precision, and
matrix dimensions. SYRK computes one triangle; the other triangle is generated
only when required by a later operation. Owned workspaces are aligned, and
wide response matrices gathered for repeated fold computations use 64-byte
column alignment. Compiled classification fitting and cross-validation
products operate on compact labels and class sums, so they do not allocate a
dense one-hot training-response matrix.

For eligible wide-response SIMPLS cross-validation, the compiled engine forms
the uncentred sample-response Gram matrix once before the fold loop. Each
training-fold Gram matrix is recovered from its principal submatrix and exactly
double-centred, avoiding a materialized centred response matrix. CPU execution
uses direct `SSYRK` or `DSYRK`; float32 CUDA uses `cublasSsyrk`; and the hybrid
Metal route uses Accelerate SYRK for this host-visible operation. The raw Gram
matrix retains only its lower triangle, and fold, output, and device workspaces
are reused across the cross-validation call.

`backend = "cuda"` uses CUDA-native routines when the package is compiled with
CUDA and a compatible NVIDIA runtime is available. After input transfer, CUDA
paths perform preprocessing, cross-products, randomized direction extraction,
component updates, low-rank prediction, and LDA scoring on the GPU for PLS-SVD,
SIMPLS, OPLS, and kernel PLS. OPLS filtering and application of the stored
orthogonal filter remain device-resident. Nonlinear kernel PLS constructs and
centers both training and test Gram matrices on the GPU. Returned arrays and R
model-object assembly are host operations. CUDA is most useful when dense
matrix work offsets transfer and kernel-launch overhead. Nonlinear kernels are
also subject to an explicit device-memory guard because their Gram matrix uses
quadratic storage in the training sample count.

`backend = "metal"` provides a float32 route for PLS-SVD, SIMPLS, OPLS, and
kernel PLS on Apple silicon unified-memory systems. It always enters the same
operation-split
algorithm rather than selecting a complete CPU or Metal estimator from the
dataset dimensions. The CPU performs preprocessing, QR and reduced
decompositions, sequential orthogonalization, deflation bookkeeping, model
assembly, LDA, and prediction. Persistent Metal workspaces perform fitting
products involving the training sample matrix. These include explicit
cross-covariance formation,
fused score/loading products, and randomized range-finder products when the
cross-covariance is represented implicitly. The smaller cross-covariance and
component-state operations remain on the CPU. This mathematical assignment is
fixed and does not switch according to dataset dimensions.
The route requires Metal support and float32 input and never silently converts
a float64 request to CPU.

## Backend and CPU Configuration

CPU is the default backend. After choosing a mathematical model, users can set
one execution backend for the R session when several calls should use the same
hardware:

```r
options(backend = "cuda")

fit <- pls(X, y, method = "simpls")    # CUDA session default
fit.cpu <- pls(X, y, backend = "cpu")  # explicit override
```

Batch jobs may use
`FASTPLS_BACKEND=cpu|cuda|metal`. Precedence is an explicit
function argument, `options(backend = ...)`, `FASTPLS_BACKEND`, and finally
CPU. For CPU execution, `options(n.cores = 4L)` requests four threads from the
linked BLAS/OpenMP runtime. Eligible matrix operations may use these threads,
but sequential PLS deflation remains serial and multicore acceleration depends
on matrix shape and the installed numerical library. An explicit
`n.cores =` argument takes precedence over the option. The first operation that
resolves an unavailable option or environment value raises an error; CPU is
never substituted. An explicitly requested `backend = "auto"` in
prediction retains model-aware prediction dispatch and errors if the fitted
model requires an unavailable accelerator.

## Supported Computational Routes

The terms in this table distinguish implementation coverage and data
residency. `Tested` denotes a route covered by fixed-seed numerical tests.
`Approximate` identifies rSVD execution. `Native` means that preprocessing,
cross-products, decomposition, component updates, prediction, and an optional
LDA head execute on the selected device after input transfer. `Unavailable`
combinations stop; they do not silently fall back to CPU.

The table uses T for tested, A for approximate, N for device-native, H for an
explicit host/device hybrid, and U for unavailable.

| Model or operation | CPU | CUDA | Metal |
|---|---:|---:|---:|
| PLS-SVD, float64 | T/A | T/A/N | U |
| SIMPLS, float64 | T/A | T/A/N | U |
| OPLS, float64 | T/A | T/A/N | U |
| Linear kernel PLS, float64 | T/A | T/A/N | U |
| Nonlinear kernel PLS, float64 | T/A | T/A/N | U |
| PLS-SVD, float32 | T/A | T/A/N | T/A/H |
| SIMPLS, float32 | T/A | T/A/N | T/A/H |
| OPLS, float32 | T/A | T/A/N | T/A/H |
| Linear kernel PLS, float32 | T/A | T/A/N | T/A/H |
| Nonlinear kernel PLS, float32 | T/A | T/A/N | T/A/H |
| Argmax PLS-DA | T | T/N | T/H |
| Latent-space LDA | T | T/N | T/H |

Float32 support is route and platform dependent, as described in the Float32
Input section. CUDA uses a device-native route, whereas Metal uses the fixed
CPU/Metal operation split. Metal float64 requests stop because the hardware
does not provide native double-precision arithmetic. Nonlinear kernel workloads that
exceed the guarded device-memory budget stop before allocation and never fall
back to another method or backend.

## Public API and Data Inputs

Users select the mathematical method and implementation through the public
functions; the internal C++, CUDA, Metal, and benchmarking helpers are not
called directly.

| Function | Purpose |
|---|---|
| `pls()` | Fit PLS models for regression or classification. |
| `predict()` | Predict from fitted `fastPLS`, OPLS, or kernel PLS models. |
| `plot()` | Plot PLS scores and optional ellipses. |
| `plot.permutation()` | Plot R2/Q2 diagnostics from a PLS permutation test. |
| `pls.single.cv()` | Select components by grouped cross-validation. |
| `pls.double.cv()` | Run nested cross-validation. |
| `evaluate()` | Evaluate classification or regression predictions. |
| `fastsvd()` | Run the stand-alone native CPU randomized SVD. |
| `fastcor()` | Compute fast Pearson-style correlations. |
| `ViP()` | Compute variable-importance-in-projection trajectories. |
| `fastPLS_blas()` | Report the CPU numerical library selected at compilation. |
| `has_cuda()` | Check whether CUDA-native fastPLS support is available. |
| `has_metal()` | Check for Apple Metal support. |

### Bioconductor Data Containers

`pls()` and `predict()` accept a `Biobase::ExpressionSet` directly. Expression
measurements are stored as variables by samples in an `ExpressionSet`, so
fastPLS transposes the assay internally to its samples-by-variables layout.
The following example predicts the case-control phenotype in the Biobase
example data while keeping the test samples separate from model fitting.

```{r biobase-expression-set}
data("sample.ExpressionSet", package = "Biobase")
y_bioc <- factor(Biobase::pData(sample.ExpressionSet)$type)

set.seed(42)
test_bioc <- sample(seq_len(ncol(sample.ExpressionSet)), 8)
fit_bioc <- pls(
    sample.ExpressionSet[, -test_bioc],
    y_bioc[-test_bioc],
    sample.ExpressionSet[, test_bioc],
    y_bioc[test_bioc],
    ncomp = 1:2,
    method = "simpls",
    backend = "cpu",
    seed = 42
)
fit_bioc$accuracy
```

### Choosing a Starting Model

| Goal | Suggested call |
|---|---|
| Fast exploratory modelling | SIMPLS with CPU rSVD |
| Repeatable CPU analysis | `backend = "cpu", seed = 1` |
| Direct cross-covariance PLS-SVD | `method = "plssvd"` |
| Remove response-unrelated structured variation | `method = "opls"` |
| Nonlinear relationships | Kernel PLS with an RBF or polynomial kernel |
| Classification: argmax PLS-DA decoding | `classifier = "argmax"` |
| Classification: latent-space discriminant analysis | `classifier = "lda"` |
| Small or moderate data | `backend = "cpu"` |
| Large dense matrices | CUDA when available |
| Apple Silicon acceleration | Metal when available |
| Float32 CPU input | `float::fl(X)`, `backend = "cpu"` |

The public classification choices are `"argmax"` and `"lda"`; regression
ignores `classifier`. LDA uses the fixed scale-normalized Cholesky fallback
sequence described above and does not expose a ridge-tuning argument.

## Classification Tasks

For classification, responses are supplied as factors. `fastPLS` handles the
PLS-DA response encoding internally and returns predicted class labels.
The `classifier` argument is used only for this type of task and selects the
classification head: `argmax` or `lda`. These options are not
regression models and are not used for numeric responses. The examples in this
section use `iris` for compact multiclass classification and
`sample.ExpressionSet` from Biobase for a biomedical two-class score plot.

```{r chunk-002}
set.seed(100)
X <- as.matrix(iris[, 1:4])
Y_cls <- iris$Species
cls_test_id <- sample(seq_len(nrow(X)), 30)
Xtrain <- X[-cls_test_id, , drop = FALSE]
Xtest <- X[cls_test_id, , drop = FALSE]
Ytrain_cls <- Y_cls[-cls_test_id]
Ytest_cls <- Y_cls[cls_test_id]
```

### Fit And Predict A Classifier

The `method` argument selects the PLS algorithm, `backend` selects the
implementation, and `classifier` selects the classification head. The default
classification head is `classifier = "argmax"`, which predicts the class with
the largest PLS-DA response score.

```{r chunk-003}
fit_cls <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    fit = TRUE,
    return_variance = FALSE,
    seed = 101
)

fit_cls$accuracy
```

Classification models can also be fitted once and predicted later. `predict()`
returns only the predicted class by default and can optionally return ranked
classes by setting `top` to a positive integer. The
ordinary prediction in `Ypred` is always the rank-1 class. When `top > 1`,
`Ypred_top` contains the ordered candidate labels for each sample: `rank1` is
the predicted class, `rank2` is the next most likely class, and so on. If
available, `Ypred_top_score` contains the corresponding class scores used to
create that ranking. For both float64 and float32 models, the default
`raw_scores = FALSE` evaluates ranked predictions in bounded row blocks and
retains only the requested ranks. Set `raw_scores = TRUE` only when the
complete class-score arrays are needed, because those arrays can be
substantially larger.

```{r chunk-004}
fit_cls_train_only <- pls(
    Xtrain,
    Ytrain_cls,
    ncomp = 1:2,
    classifier = "lda",
    fit = TRUE,
    return_variance = FALSE,
    seed = 101
)

pred_cls_later <- predict(
    fit_cls_train_only,
    Xtest,
    Ytest = Ytest_cls,
    top = 2,
    raw_scores = TRUE
)

pred_cls_later$accuracy
pred_cls_later$metrics$metrics
head(pred_cls_later$Ypred_top[["ncomp=2"]])
```

### Evaluate Classification Predictions

Use `evaluate()` to summarize predicted class labels, ranked labels,
class-score matrices, or a complete object returned by `predict()`. The task
and requested rank are inferred from these inputs.
For classification, `lift_accuracy` is accuracy divided by the
`no_information_rate`, the accuracy obtained by always predicting the most
frequent observed class. Values above one therefore improve on this simple
majority-class baseline.
For classification, the complete output includes global metrics, per-class
metrics, and the confusion matrix.

```{r chunk-005}
eval_cls_path <- evaluate(
    observed = Ytest_cls,
    predicted = pred_cls_later
)

eval_cls <- eval_cls_path$by_component[["ncomp=2"]]
eval_cls
```

The same complete evaluation is also stored automatically in a fitted PLS
object when observed test responses are supplied. Results are grouped first by
data role and then by component count.

```{r chunk-005a}
fit_cls$metrics$test[["ncomp=2"]]$metrics
```

Rows of the confusion matrix are predicted labels and columns are observed
labels. The confusion matrix is returned as an ordinary R table.

```{r chunk-006}
eval_cls$confusion
```

<div style="page-break-before: always;"></div>

When class-score matrices are available, `evaluate()` can also report top-k
accuracy. Top-k accuracy asks whether the true class appears anywhere among the
first `k` ranked labels. Thus top-1 accuracy is ordinary classification
accuracy,
while top-5 accuracy gives credit when the correct class appears among the five
highest-scoring alternatives.

```{r chunk-007}
score_last <- pred_cls_later$LDA_scores[
    , , dim(pred_cls_later$LDA_scores)[3L]
]

evaluate(
    observed = Ytest_cls,
    predicted = score_last
)
```

### Classification Heads

The same `pls()` interface exposes two classification-specific heads for
factor responses: argmax PLS-DA and latent-space LDA. They are
decoders applied after the PLS model has produced class-response scores or
latent scores. For regression, leave `classifier` at its default; numeric
responses are predicted directly as continuous values.

```{r chunk-008}
fit_cls_plssvd <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "plssvd",
    seed = 100
)

head(fit_cls_plssvd$Ypred)

evaluate(
    observed = Ytest_cls,
    predicted = fit_cls_plssvd$Ypred[["ncomp=2"]]
)$confusion
```

For PLS-DA with an LDA prediction head, use `classifier = "lda"`. On systems
with GPU support, the backend is selected through `backend = "cuda"` or
`backend = "metal"` where available:

```{r chunk-009, eval = FALSE}
fit_cls_lda_gpu <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "plssvd",
    backend = "cuda",
    classifier = "lda"
)
```

An unavailable CUDA or Metal selection raises an error; fastPLS never changes
the request silently to CPU. To run on the CPU, select it explicitly:

```{r chunk-010}
fit_cls_lda_cpu <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "plssvd",
    backend = "cpu",
    seed = 100,
    classifier = "lda"
)

head(fit_cls_lda_cpu$Ypred)
```

### Kernel PLS For Classification

Kernel PLS changes the representation of the samples before the inner PLS fit.
The `linear` kernel is equivalent to an ordinary inner-product representation
    and
is useful as a fast baseline. The `rbf` kernel uses a radial-basis similarity;
the `poly` kernel uses polynomial interactions among features.

```{r chunk-012}
kernel_fits <- lapply(c("linear", "rbf", "poly"), function(k) {
    pls(
    Xtrain, Ytrain_cls, Xtest, Ytest_cls,
    ncomp = 1:2,
    method = "kernelpls",
    kernel = k,
    degree = 2,
    seed = 102
    )
})
names(kernel_fits) <- c("linear", "rbf", "poly")

kernel_accuracy <- vapply(kernel_fits, function(fit) {
    mean(fit$Ypred[["ncomp=2"]] == Ytest_cls)
}, numeric(1))
kernel_accuracy
```

### Classification Score Plots

`plot()` can visualize stored PLS score maps. Refit with `fit = TRUE` to store
training scores, or predict with `proj = TRUE` to store test scores. Ellipses
    can
be ordinary confidence ellipses or Hotelling T2 ellipses.

```{r chunk-013, fig.width = 5, fig.height = 4}
plot(
    fit_cls,
    groups = Ytrain_cls,
    ellipse = TRUE,
    ellipse.type = "confidence"
)
```

The next example uses the two-class gene-expression data from the Biobase
`sample.ExpressionSet` object introduced above. The same `plot()` call works for
`method = "plssvd"`, `"simpls"`, `"opls"`, and `"kernelpls"`.

```{r biobase-binary-pls-plots, fig.width = 5.5, fig.height = 4.5}
X_two <- t(Biobase::exprs(sample.ExpressionSet))
y_two <- factor(Biobase::pData(sample.ExpressionSet)$type)
train_two <- setdiff(seq_along(y_two), test_bioc)
fit_two <- pls(
    X_two[train_two, , drop = FALSE], y_two[train_two],
    X_two[test_bioc, , drop = FALSE], y_two[test_bioc],
    ncomp = 1:2, fit = TRUE, proj = TRUE, seed = 200
)

plot(fit_two, score.set = "train", groups = y_two[train_two],
        ellipse = TRUE, main = "Biobase expression data")
```

The second score-plot example fits OPLS to the three-class `iris` split defined
at the beginning of this section, with two orthogonal components (`north = 2`).

```{r iris-three-class-opls, fig.width = 7.5, fig.height = 4}
opls_three <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "opls",
    fit = TRUE,
    proj = TRUE,
    north = 2L,
    seed = 201
)

old_par <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 0.8))
plot(
    opls_three,
    score.set = "train",
    groups = Ytrain_cls,
    ellipse = TRUE,
    ellipse.type = "hotelling",
    main = "OPLS training, north = 2"
)
plot(
    opls_three,
    score.set = "test",
    groups = opls_three$Ypred[["ncomp=2"]],
    xlim = c(-0.16, 0.42),
    main = "OPLS test prediction"
)
par(old_par)
```

## Regression Tasks

For regression, responses are supplied as numeric vectors or matrices. The
examples in this section use `mtcars` to demonstrate univariate regression and
OPLS. Regression does not use `classifier = "argmax"` or `classifier = "lda"`:
those are classification heads for factor responses.
For numeric responses, `pls()` returns continuous predictions and regression
metrics such as `R2Y`, `Q2Y`, and `RMSD`.

```{r chunk-016}
set.seed(100)
Xreg <- as.matrix(mtcars[, c("disp", "hp", "wt", "qsec", "drat")])
Y_reg <- matrix(mtcars$mpg, ncol = 1)
reg_test_id <- sample(seq_len(nrow(Xreg)), 8)
Xreg_train <- Xreg[-reg_test_id, , drop = FALSE]
Xreg_test <- Xreg[reg_test_id, , drop = FALSE]
Ytrain_reg <- Y_reg[-reg_test_id, , drop = FALSE]
Ytest_reg <- Y_reg[reg_test_id, , drop = FALSE]
```

### Fit And Predict A Regression Model

The simplest workflow is to provide both the training data and an independent
test set directly to `pls()`. In this case the fitted object already contains
test-set predictions and, if `Ytest` is supplied, predictive metrics.

```{r chunk-017}
fit_reg <- pls(
    Xreg_train,
    Ytrain_reg,
    Xreg_test,
    Ytest_reg,
    ncomp = 1:3,
    fit = TRUE,
    return_variance = FALSE
)

fit_reg$Q2Y
```

### Float32 Input

`pls()` also accepts predictor and response matrices from the `float` package.
Supplying a `float::float32` object requests single-precision execution. CUDA
keeps PLS-SVD, SIMPLS, OPLS, nonlinear kernel construction, prediction, and LDA
on the selected GPU. Metal accepts float32 only and uses the fixed CPU/Metal
operation split described above; CUDA supports float32 and float64. Nonlinear
kernel PLS requires an explicit `n` by `n` Gram matrix, so the package checks
the estimated live device or unified-memory storage before fitting.
Extreme-response and nonlinear-kernel routes should be checked against float64
when the scientific conclusion is sensitive to small numerical differences.

Float64 remains the baseline precision. Float32 approximately halves raw
matrix storage, but paired tests show that runtime, incremental host RSS, and
device memory may increase or decrease depending on the route. The package
therefore warns once for measured-risk regimes. This
includes precision-sensitive SIMPLS and linear kernel-PLS classification,
nonlinear kernels and multivariate regression
with at least 10,000 response columns and 50 components. In that last regime,
runtime, memory, and numerical behavior must be checked against float64 before
the float32 result is used for scientific interpretation. Unsupported
combinations stop before allocation and are never silently promoted to
float64.

On Windows, the standard R toolchain does not expose the single-precision
BLAS/LAPACK symbols used by the Unix-like compiled kernels. The Windows CPU
route therefore combines float-package rSVD and Cholesky operations with
portable C++ float kernels. It supports PLS-SVD, SIMPLS, OPLS, linear and
nonlinear kernel PLS, argmax, and latent-space LDA when `backend = "cpu"` and
rSVD is selected automatically. Model factors, LDA buffers, kernel matrices, and
predictions remain float32, but this portable route can be slower than the
native Unix-like implementation. Windows float32 accelerator requests
stop with an unsupported-combination error rather than silently
converting data to float64.

```{r chunk-018}
Xreg32 <- float::fl(as.matrix(Xreg_train))
Yreg32 <- float::fl(matrix(Ytrain_reg, ncol = 1))
fit_reg32 <- pls(
    Xreg32,
    Yreg32,
    float::fl(as.matrix(Xreg_test)),
    float::fl(matrix(Ytest_reg, ncol = 1)),
    ncomp = 1:2
)
fit_reg32$Q2Y
```

For standard double-precision regression, `Ypred` is a numeric prediction array
with one slice for each requested number of components. For float32 input,
`Ypred` is kept as a named list of `float::float32` prediction matrices. In the
standard example below, the last array slice corresponds to the largest
requested component count (`ncomp = 3`). The metric vectors are named by
component count, for example `fit_reg$Q2Y["ncomp=3"]`.

```{r chunk-019, fig.width=4.8, fig.height=4}
reg_component <- dim(fit_reg$Ypred)[3L]
pred_mpg <- fit_reg$Ypred[, , reg_component]
plot(
    Ytest_reg,
    pred_mpg,
    pch = 21,
    bg = "#4E79A7",
    col = "black",
    xlab = "Observed mpg",
    ylab = "Predicted mpg",
    main = "Regression: observed vs predicted"
)
abline(0, 1, col = "#D55E00", lwd = 2)
```

The alternative workflow is to fit the model once without a test set, then call
`predict()` later. This is useful when the same model must be applied to several
independent datasets. `predict()` automatically applies the centering/scaling
stored in the fitted object. When `Ytest` is supplied, it also passes the
predictions to `evaluate()` and stores the complete evaluation under `metrics`.

```{r chunk-020}
fit_reg_train_only <- pls(
    Xreg_train,
    Ytrain_reg,
    ncomp = 1:3,
    fit = TRUE,
    return_variance = FALSE
)

pred_reg_later <- predict(
    fit_reg_train_only,
    Xreg_test,
    Ytest = Ytest_reg,
    proj = TRUE
)

pred_reg_later$Q2Y
pred_reg_later$metrics$metrics
head(pred_reg_later$Ttest)
```

### Evaluate Regression Predictions

For numeric regression, `evaluate()` reports R2, Q2, RMSD/RMSE, MAE, bias,
median relative error percentage, RPD, and correlations. If `ytrain` is
supplied, Q2 is calculated relative to the training-set response mean, which
is the preferred setting for independent test-set evaluation. For a response
matrix, each response column is centered on its corresponding training mean
before the denominator sums are aggregated. If `ytrain` is omitted, Q2 is
returned as `NA`; it is not silently replaced by an R2 calculation.

For wide multivariate responses, `pls()`, `pls.single.cv()`, and
`pls.double.cv()` return aggregate evaluation metrics by default. Set
`bycol = TRUE` to also calculate the full response-wise metric table; this is
useful for inspecting individual spectral bins, but can be expensive for NMR
data with many response columns.

```{r chunk-021}
eval_reg <- evaluate(
    observed = Ytest_reg,
    predicted = pred_mpg,
    ytrain = Ytrain_reg
)

eval_reg$task
eval_reg$metrics
lapply(eval_reg$metric_definitions, strwrap, width = 56)
eval_reg$per_response
```

### OPLS For Regression

OPLS is accessed through the same `pls()` function with `method = "opls"`. It is
often used to separate predictive variation from response-orthogonal variation.
`ncomp` counts predictive components, separately from the orthogonal components
removed by `north`. Each removed direction reduces the available predictor
rank. For example, with eight independent predictors and one removed direction,
at most seven predictive components remain. A request exceeding the remaining
rank raises an error; the same restriction applies within each CV training fold.

```{r chunk-022}
fit_opls <- pls(
    Xreg_train,
    Ytrain_reg,
    Xreg_test,
    Ytest_reg,
    ncomp = 1:2,
    method = "opls",
    seed = 101
)

class(fit_opls)
fit_opls$Q2Y
```

## Cross-Validation

`fastPLS` provides two cross-validation helpers. Use `pls.single.cv()` for
grouped k-fold or leave-one-group-out validation; pass a scalar `ncomp` for a
fixed-component CV or a vector of candidates when the number of components
should be selected from a grid. Use `pls.double.cv()` for nested validation,
where an inner CV chooses the number of components and an outer CV estimates
predictive performance. Both helpers support regression and classification and
dispatch to compiled CPU or accelerator routes when available. CUDA SIMPLS CV
keeps task data and fold workspaces on the device. CUDA library handles are
created once per CV call and reused across folds rather than reconstructed for
every fit. CUDA PLS-SVD CV also uses a resident route for large multivariate
regression responses, while
classification and smaller-response tasks retain the faster fold-local CUDA
route. This choice is automatic and does not change fold assignment,
fold-specific preprocessing, requested components, or prediction semantics.
For large multivariate responses, predictor and response marginal moments are
computed once. Each training fold obtains its own moments by subtracting the
held-out contribution before fold-specific centering and scaling. The large
cross-covariance remains matrix-free, so this reuse does not introduce a dense
predictor-by-response cache. When its cost and storage model is favorable, the
compiled CPU engine can also form a bounded response Gram matrix once for
wide-response SIMPLS problems and extract an exactly double-centred training
submatrix for each fold. The Gram kernel is selected between GEMM and SYRK by
platform, precision, and matrix shape; a SYRK route retains one triangle until
a complete fold matrix is required. Fold-specific
preprocessing, seeds, and model fitting remain unchanged. Aggregate regression
metrics are computed directly from the native out-of-fold predictions before
those predictions are formatted for R.

### Ordinary k-fold CV

In ordinary k-fold CV, samples are split directly into `kfold` folds. The
example below validates a fixed two-component SIMPLS classifier with five folds.

```{r chunk-023}
cv_kfold <- pls.single.cv(
    Xdata = Xtrain,
    Ydata = Ytrain_cls,
    ncomp = 2,
    kfold = 5,
    seed = 103
)

cv_kfold$metrics$cross_validated[["ncomp=2"]]$metrics
```

`metrics$cross_validated` contains the complete `evaluate()` output for each
tested component count, while `selection_metrics` is the compact internal table
used to select the best setting. When `fit = TRUE`, `metrics$fitted` evaluates
the corresponding model fitted to the full dataset.

### Grouped k-fold CV with `constrain`

The `constrain` argument controls grouped splitting. It is a vector with one
entry per sample; samples with the same value are assigned to the same fold. In
practice, this prevents leakage when multiple rows come from the same patient,
subject, batch, or technical replicate. For example, if two spectra come from
the same patient, giving them the same patient identifier in `constrain` ensures
that both spectra are placed either in the training set or in the test set,
never one in each.

```{r chunk-024}
patient_id <- rep(
    seq_len(ceiling(nrow(Xtrain) / 2)), each = 2
)[seq_len(nrow(Xtrain))]

cv_grouped <- pls.single.cv(
    Xdata = Xtrain,
    Ydata = Ytrain_cls,
    constrain = patient_id,
    ncomp = 2,
    kfold = 4,
    seed = 104
)

c(n_folds = length(unique(cv_grouped$fold)),
    n_patient_groups = length(unique(patient_id)))
```

### Leave-one-out CV and leave-one-group-out CV

Leave-one-out CV is requested with `kfold = "loocv"`. When `constrain` is not
supplied, each sample is held out once. When `constrain` is supplied, LOOCV
becomes leave-one-constraint-group-out CV, so a whole patient, subject, batch,
or replicate group is held out together at each iteration. Numeric `kfold`
values greater than or equal to the number of constraint groups are also treated
as leave-one-group-out CV.

```{r chunk-025}
cv_loocv <- pls.single.cv(
    Xdata = Xtrain,
    Ydata = Ytrain_cls,
    constrain = patient_id,
    ncomp = 1,
    kfold = "loocv",
    seed = 105
)

c(n_loocv_folds = length(unique(cv_loocv$fold)),
    n_patient_groups = length(unique(patient_id)))
```

### Component and hyperparameter optimization CV

`pls.single.cv()` repeats the same CV splitting strategy over several candidate
component counts and returns the best value according to the predictive metric:
accuracy for classification and RMSD for regression. Fold-aware Q2Y,
full-data fitted R2Y, balanced accuracy, and the other task-specific metrics
documented below can instead be requested explicitly through `selection`.
Predictive arguments can
also be supplied as vectors to tune a compact grid. Fold-control arguments such
as `kfold` remain single settings for the whole CV run.

For LDA classification, a rare class may be absent from an inner training
fold. fastPLS fits LDA to the classes represented in that fold and maps the
predictions back to the original factor levels. Held-out observations from an
absent class are retained in the accuracy and balanced-accuracy calculations,
although that class cannot be predicted by a model that did not observe it.
When otherwise identical argmax and LDA configurations are requested together,
fastPLS calculates the PLS component path and fold projection once. The two
classification heads retain separate predictions, metric paths, and selected
component counts. For sufficiently tall SIMPLS classification problems,
full-data predictor and class moments are also computed once. Each training
fold obtains its moments by subtracting the held-out contribution before
applying fold-specific centering and scaling; the PLS model, LDA model, and
predictions are still fitted independently within every fold. This avoids
repeatedly materializing large training-score matrices without leaking
held-out information into model fitting. The same sufficient statistics can
assemble eligible PLS-SVD folds through the fold-specific predictor Gram and
class cross-product, without materializing the training-score matrix. On the
Metal route, small reduced products are evaluated with Accelerate while the
large sample-matrix products assigned to Metal remain on the GPU.

```{r chunk-026}
cv_opt <- pls.single.cv(
    Xdata = Xreg_train,
    Ydata = Ytrain_reg,
    ncomp = 1:3,
    kfold = 5
)

cv_opt$best_ncomp
```

### Interpreting `R2Y`, `Q2Y`, and `RMSD`

For regression models, `R2Y`, `Q2Y`, and `RMSD` answer different questions and
should not usually be identical. In `pls()`, `R2Y` is training-set R2 and
independent-test `Q2Y` uses the training-response mean in its denominator. For
multivariate responses, each response is centered separately before the sums
of squares are aggregated. In `pls.single.cv()`, each held-out prediction is
evaluated relative to the corresponding fold-training response means.
`pls.double.cv()` applies the same rule to the outer folds. `RMSD` is also
calculated
from held-out predictions and is reported on the response scale, so lower values
are better. `R2Y` is a training-set explained-variance estimate from one
additional model fitted on the full dataset; set `fit = FALSE` to skip
this extra fit when only cross-validated performance is needed. For
classification, `Q2Y` is calculated from held-out dummy-coded PLS-DA response
scores using fold-training class proportions, `accuracy` reports decoded-label
accuracy, and `R2Y` is calculated from the full-data PLS-DA fit on the
dummy-coded response scores. Dummy-response `Q2Y` and `R2Y` are not
classification accuracy. The exact convention used by each function is also
reported in its `metrics$definitions` element. `RMSD` is not used for
classification.

Use `selection = "R2Y"` to select from the full-data fitted-response path or
`selection = "Q2Y"` to select from out-of-fold predictions standardized by
each fold's training-response mean. Selecting R2Y automatically enables and
returns the fitted path even if `fit = FALSE`; because it is a training
criterion, Q2Y or another held-out metric is usually preferable for choosing
model complexity. The former names `"r2"` and `"q2"` are rejected because they
did not identify these two different quantities unambiguously.

Classification settings can be selected by `accuracy`, `balanced_accuracy`,
`lift_accuracy`, `macro_precision`, `macro_recall`, `macro_f1`, `kappa`, R2Y,
or Q2Y. Use balanced accuracy when unequal class frequencies make
majority-class accuracy misleading. Regression settings can be selected by
R2Y, Q2Y, `RMSD`, `MAE`, `MAPE_percent`, `RPD`, `Pearson_r`, or
`Spearman_r`. These additional aggregate metrics are useful for
high-dimensional multivariate responses such as spectra. RMSD, MAE,
and MAPE_percent are minimized; the other criteria are maximized. Signed bias
and signed `MRE_percent` remain available from `evaluate()` but are not tuning
criteria because they have no unambiguous one-sided optimization direction.
The definitions match `evaluate()`, and an incompatible choice such as
classification accuracy for a regression response raises an error before
model fitting.

```{r chunk-027}
data.frame(
    ncomp = cv_opt$ncomp,
    training_R2Y = round(cv_opt$R2Y, 3),
    heldout_Q2Y = round(cv_opt$Q2Y, 3),
    heldout_RMSD = round(cv_opt$RMSD, 3)
)
```

### Permutation-test p-values

`fastPLS` provides two permutation-test procedures. In `pls()`, the permutation
test is a single train/test procedure: the rows of `Xtrain` are randomly
permuted, the model is refitted, and the permuted test-set `Q2Y` values are
compared with the observed `Q2Y` values component by component. The returned
`pval` is the corrected Monte Carlo upper-tail value
`(b + 1) / (B + 1)`, where `b` counts successful null fits at least as extreme
as observed and `B` counts successful null fits. It can therefore never be zero.
Because `pls()` has no grouping argument, one training row is the permutation
unit. The full permutation table is stored in `permutation` and can
be visualized with `plot.permutation()`, where the x-axis is the correlation
between the original and permuted response structure and the y-axis shows R2
and Q2.

```{r chunk-028}
perm_fit <- pls(
    Xtrain = Xreg_train,
    Ytrain = Ytrain_reg,
    Xtest = Xreg_test,
    Ytest = Ytest_reg,
    ncomp = 2,
    fit = TRUE,
    perm.test = TRUE,
    return_variance = FALSE,
    seed = 108
)

perm_fit$pval
plot.permutation(perm_fit, ncomp = 2)
```

In `pls.double.cv()`, the permutation test repeats the complete nested
cross-validation workflow. Independent rows are exchanged individually. When
`constrain` identifies repeated observations, complete constraint blocks are
exchanged only between groups with the same number of rows. Thus group sizes,
within-group response structure, and class frequencies are preserved exactly.
The observed and permuted analyses use identical outer and inner folds and the
same randomized-SVD seeds, so null variation reflects the exchange operation
rather than new folds or sketches. Model selection and permutation inference
use the same `selection`. For imbalanced
classification, `selection = "balanced_accuracy"` therefore selects the
model by mean class-specific recall and tests that statistic against its
permutation distribution; it does not substitute dummy-response Q2. The output
records the statistic in `permutation_metric`, `permutation_observed`, and
`permutation_sampled`. Larger predictive metrics use the upper permutation tail,
whereas losses such as RMSD use the lower tail. Both tails use
`(b + 1) / (B + 1)`. Failed null fits are stored in `permutation_errors`,
    omitted
from `B`, and summarized by `permutation_completed` and `permutation_failed`.

```{r chunk-029}
dcv_perm <- pls.double.cv(
    Xdata = Xreg_train[1:20, ],
    Ydata = Ytrain_reg[1:20],
    ncomp = 1:2,
    kfold_inner = 2,
    kfold_outer = 2,
    perm.test = TRUE,
    seed = 109
)

data.frame(
    permutation_metric = dcv_perm$permutation_metric,
    observed = dcv_perm$permutation_observed,
    p_value = dcv_perm$p.value,
    completed = dcv_perm$permutation_completed,
    failed = dcv_perm$permutation_failed
)
```

For an imbalanced classification analysis, the corresponding call is:

```{r balanced-permutation-example, eval=FALSE}
dcv_balanced <- pls.double.cv(
    Xdata = Xtrain,
    Ydata = Ytrain_cls,
    ncomp = 1:5,
    kfold_inner = 5,
    kfold_outer = 5,
    constrain = patient_id,
    selection = "balanced_accuracy",
    perm.test = TRUE,
    times = 100,
    seed = 109
)
dcv_balanced$balanced_accuracy
dcv_balanced$metrics$permutation
```

The selected CV object can then be passed directly to `pls()`. The final model
is refitted on the full training set using the selected component count and
selected tuning settings, then applied to the independent test set.

```{r chunk-030}
cv_select <- pls.single.cv(
    Xdata = Xtrain,
    Ydata = Ytrain_cls,
    ncomp = 1:3,
    kfold = 5,
    seed = 106
)

fit_selected <- pls(
    cv_select,
    Xtest = Xtest,
    Ytest = Ytest_cls,
    return_variance = FALSE
)

data.frame(
    best_ncomp = cv_select$best_ncomp,
    test_accuracy = mean(fit_selected$Ypred[[1]] == Ytest_cls)
)
```

For example, `kernelpls` can select the best combination of component count and
kernel setting. The selected values are returned in `best_parameters`, and the
same CV object can be passed to `pls()` for the final refit.

```{r chunk-031}
cv_kernel <- pls.single.cv(
    Xdata = Xtrain,
    Ydata = Ytrain_cls,
    ncomp = 1:3,
    kfold = 5,
    method = "kernelpls",
    kernel = c("linear", "rbf"),
    gamma = c(0.1, 1),
    seed = 107
)

fit_kernel <- pls(
    cv_kernel,
    Xtest = Xtest,
    Ytest = Ytest_cls,
    return_variance = FALSE
)

data.frame(
    best_ncomp = cv_kernel$best_parameters$ncomp,
    best_kernel = cv_kernel$best_parameters$kernel,
    test_accuracy = mean(fit_kernel$Ypred[[1]] == Ytest_cls)
)
```

### Nested or double CV

Double cross-validation is a nested validation design for separating model
optimization from the final estimate of predictive performance. This separation
is especially important for PLS-DA because the number of latent variables and
other modelling choices can otherwise be tuned on the same samples used to
report performance, producing optimistic accuracy or Q2 estimates. In the
terminology of Szymanska et al. (2012), the inner CV loop (`CV1`) is used to
optimize model complexity, such as the number of latent variables, whereas the
outer CV loop (`CV2`) holds out samples that are not used during optimization
    and
therefore provides the performance estimate of the complete modelling strategy.
In `fastPLS`, `pls.double.cv()` follows this idea using a reproducible grouped-
fold plan constructed in R. CPU and Metal use a compiled coordinator for inner
selection, outer refitting, prediction, and metric accumulation. CUDA uses an R
coordinator around native CUDA single-CV and outer-fit kernels. Both
`kfold_inner` and `kfold_outer` can be ordinary fold counts or `"loocv"`, and
both respect `constrain`.

```{r chunk-032}
cv_double <- pls.double.cv(
    Xdata = Xtrain,
    Ydata = Ytrain_cls,
    constrain = patient_id,
    ncomp = 1:2,
    kfold_inner = 3,
    kfold_outer = 3,
    seed = 104
)

data.frame(
    selected_ncomp_mode = cv_double$bcomp,
    outer_metric = cv_double$metric_name,
    outer_accuracy = cv_double$accuracy,
    outer_Q2Y = cv_double$Q2Y,
    outer_R2Y = cv_double$R2Y
)
```

The same interface can tune more than the component count. In this example the
inner loop chooses the best combination of `ncomp`, `kernel`, and `gamma` for
`kernelpls`. The outer loop then uses the selected combination for each held-out
outer fold.

```{r chunk-033}
cv_double_kernel <- pls.double.cv(
    Xdata = Xtrain,
    Ydata = Ytrain_cls,
    constrain = patient_id,
    ncomp = 1:2,
    kfold_inner = 3,
    kfold_outer = 3,
    method = "kernelpls",
    kernel = c("linear", "rbf"),
    gamma = c(0.1, 1),
    seed = 108
)

cv_double_kernel$results[[1]]$best_parameters
```

The most commonly used `pls.double.cv()` output fields are:

| Field | Meaning |
|---|---|
| `bcomp` | Most frequently selected number of components. |
| `accuracy`, `Q2Y`, `RMSD` | Outer-CV predictive performance. |
| `Ypred` | Final cross-validated prediction for each sample. |
| `conf` | Classification confusion matrix. |
| `metrics` | Detailed outer-CV evaluation results. |
| `results` | Detailed per-run and per-fold information. |

Important elements returned by `pls.double.cv()` are:

- `results`: one entry per repeated outer CV run. Each entry stores `Ypred` and
    `pred` for that run, the outer `fold` assignment, the `best_ncomp` selected
    inside each outer fold, the full `best_parameters` selected inside each
        outer
    fold, the complete inner-CV objects in `inner`, the run-level `metric_name`
    and `metric_value`, and the fitted `backend` and `method`.
- `Ypred`: the final cross-validated prediction for each sample. For
    classification, repeated runs are combined by voting; for regression,
    predictions are averaged across repeated runs.
- `acc_tot`: classification-only text summary of the total number and
    percentage of correctly classified samples.
- `conf`: classification-only confusion matrix. Entries are printed as counts
    and column percentages so that class-wise errors can be inspected.
- `vote_counts`: classification-only matrix with one row per sample and one
    column per class, showing how many repeated outer-CV runs voted for each
    class.
- `accuracy`, `Q2Y`, `RMSD`, and `R2Y`: one value per repeated outer CV run.
    For classification, `accuracy` reports decoded-label accuracy, `Q2Y` reports
    held-out Q2 on dummy-coded PLS-DA response scores, `R2Y` reports the mean
    training-fit R2 of the selected outer-fold PLS-DA models, and `RMSD` is not
    used. For regression, `Q2Y` reports held-out Q2, `RMSD` reports held-out
    RMSD, and `R2Y` reports the mean training-fit R2 of the selected outer-fold
    models.
- `metric_name`: the metric used for inner/outer model selection. It is based
    on held-out predictions except when `selection = "R2Y"` explicitly requests
    the fitted-response criterion.
- `medianQ2Y`, `CI95Q2Y`, `medianR2Y`, `CI95R2Y`, `medianRMSD`, and
    `CI95RMSD`: summaries across repeated outer CV runs, returned only when
    `runn > 1`. They are omitted for the default `runn = 1` output.
- `bcomp`: the most frequently selected number of components across all outer
    folds and repeated runs.
- `backend` and `method`: the default backend and PLS method used by the call.
    If vector-valued methods or backends are tuned in the inner loop, the
        selected
    fold-level values are stored in `results[[run]]$best_parameters`.
- `selection_metric`: the criterion used by the inner CV loop. The default
    `"auto"` means accuracy for classification and RMSD-style prediction error
    for multivariate regression. All task-compatible selection criteria listed
    above can also drive the nested permutation test.

## SVD Utility

`fastsvd()` provides direct access to the native CPU randomized SVD used by
PLS. It returns the left singular vectors (`u`), singular values (`d`), and
right singular vectors (`v`) for users who want a truncated decomposition
outside a PLS model. A standard R numeric matrix selects float64 computation.
A matrix created with `float::fl()` selects the native float32 path
automatically; its left and right singular vectors remain in float32 format.
The function exposes no solver selector because rSVD is its only algorithm.
Stand-alone CUDA and Metal SVD routes are not exposed; their reduced
decompositions are not device-native for every matrix shape.

```{r chunk-034}
s64 <- fastsvd(Xtrain, ncomp = 3, seed = 104)
s32 <- fastsvd(float::fl(Xtrain), ncomp = 3, seed = 104)
names(s64)
s64$d
inherits(s32$u, "float32")
```

## CUDA and Metal Availability

CUDA and Apple Metal are optional. Use `has_cuda()` before selecting
`backend = "cuda"` and `has_metal()` before selecting `backend = "metal"`.
An unavailable accelerator request stops with an informative error and is not
redirected to CPU.

```{r chunk-038}
has_cuda()
has_metal()
```

```{r chunk-039, eval = FALSE}
if (has_cuda()) {
    fit_gpu <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    backend = "cuda"
    )

    fit_gpu_lda <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "plssvd",
    backend = "cuda",
    classifier = "lda"
    )
}
```

```{r chunk-040, eval = FALSE}
if (has_metal()) {
    Xtrain_metal <- float::fl(as.matrix(Xtrain))
    Xtest_metal <- float::fl(as.matrix(Xtest))
    fit_metal <- pls(
        Xtrain_metal,
        Ytrain_cls,
        Xtest_metal,
        Ytest_cls,
        ncomp = 1:2,
        backend = "metal"
    )
}
```

## Helper Functions

`fastcor()` computes fast Pearson-style correlations. `ViP()` returns variable
importance in projection trajectories for fitted PLS models, following the
standard VIP interpretation used for PLS variable ranking (Wold, Sjostrom, and
Eriksson, 2001; Chong and Jun, 2005). VIP is most useful when the columns of
`X` are interpretable predictors, such as genes, metabolites, spectral bins, or
clinical variables; larger values indicate stronger contribution to the fitted
latent predictive model. Their matrix calculations, like the metrics returned
by `evaluate()`, are implemented in the dependency-free C++ core and are
independent of CUDA or Metal availability; the R functions validate inputs and
format the returned objects.

```{r chunk-041}
C <- fastcor(Xtrain, byrow = FALSE, diag = FALSE)
dim(C)
```

```{r chunk-042}
vip <- ViP(fit_reg)
dim(vip)
```

## Detailed Algorithmic Notes

### Singular-vector backends

The PLS algorithms in `fastPLS` are separated from the inner singular-vector
backend. The compiled implementation uses native randomized SVD, estimating
the leading singular triplets needed by the requested PLS components.
The randomized SVD backend
follows the randomized range-finder framework described by Halko, Martinsson,
and Tropp (2011): it draws a Gaussian test matrix, forms a low-dimensional
sketch of the target matrix, optionally applies power iterations,
orthonormalizes the sketch, and solves the final SVD in the small projected
space. In `pls()` and the cross-validation helpers, these controls are supplied
directly through `...`; in stand-alone SVD calls they are supplied directly to
`fastsvd()`.

Standalone accelerator `fastsvd()` routes are unavailable because their
reduced QR and SVD stages are not fully device-native for every matrix shape.
The function raises an error rather than returning a partial result. CUDA
acceleration remains device resident within PLS, while Metal acceleration uses
the fixed CPU/Metal operation split described below.

The PLS Metal backend, `pls(..., backend = "metal")`, supports float32 PLS-SVD,
SIMPLS, OPLS, and kernel PLS with Metal Performance Shaders. It returns
host-accessible model factors and always uses a
fixed mathematical division: Metal performs fitting products involving the
training sample matrix, including explicit cross-covariance formation, fused
score/loading products, and implicit randomized range-finder products; CPU code performs the
reduced decomposition, sequential PLS algebra, LDA, prediction, and model
assembly. Dataset dimensions never select a complete CPU or Metal route. The
executed path, storage decision, and randomized controls are reported in
`diagnostics$metal_operation_split`. Ordinary R double matrices produce an
explicit error rather than silently falling back to the CPU; use `float::fl()`
to request Metal float32 execution. Repeated fits in one R session reuse
initialized device and pipeline state, whereas an isolated first call includes
that setup cost.

For CUDA builds, GPU execution is selected with `backend = "cuda"`, so the
numerical fitting and prediction stages are performed by resident GPU-native
PLS routines after input transfer. When the
cross-covariance \(S = X^T Y\) would be large, selected backends can use the
matrix-free `xprod` route. In that route, the code avoids explicitly storing
`S` and evaluates the products required by the SVD through identities such as
\(Sv = X^T(Yv)\) and \(S^Tu = Y^T(Xu)\). For classification problems with many
classes, label-aware products can avoid materializing a dense one-hot response
matrix. Compact prediction instead retains the PLS latent factors and evaluates
only the requested test-score blocks, avoiding dense coefficient and prediction
paths when they are unnecessary.
fastPLS chooses the matrix-free route automatically only when the
cross-product is large enough to justify it.

For `pls.single.cv()`, seeded grouped fold construction and the numerical CV
path are compiled. For `pls.double.cv()`, CPU and Metal nested loops are
compiled after R constructs the fixed fold plan; CUDA retains R-level nested
coordination around native CUDA kernels. Final R-object assembly remains
host-side. CUDA
SIMPLS CV uploads the task once and
keeps fold gathering, training-only standardization, fitting, prediction, and
metric reduction on the device. CUDA PLS-SVD uses the same resident strategy
for large multivariate regression responses; classification and smaller
responses use a fold-local CUDA route because it is faster for those workloads.
Metal applies the fixed CPU/Metal operation split used by `pls()`. Unsupported
accelerator combinations stop rather than falling back to a CPU fit.

### PLS-SVD

`method = "plssvd"` implements the direct cross-covariance SVD formulation of
PLS, within the same broad PLS regression and PLS-DA framework described by
Wold, Sjostrom, and Eriksson (2001), Barker and Rayens (2003), and Boulesteix
and Strimmer (2007). After preprocessing, the algorithm computes the dominant
singular subspace of \(X^T Y\) and reuses that subspace to generate the
requested component path.
The maximum meaningful number of PLS-SVD components is capped by
`min(nrow(X), ncol(X), ncol(Y))`; if a larger value is requested, the package
uses the valid internal rank and records the capped component count.
For an implicit cross-covariance, numerical rank is determined on the same
singular-value scale as the direct decomposition. When a reduced Gram shortcut
does not contain the complete requested retained rank, PLS-SVD uses the direct
reduced decomposition rather than silently returning fewer directions.

The main innovation relative to a simple textbook PLS-SVD implementation is the
execution strategy. fastPLS combines randomized SVD, compact
low-rank prediction factors, CUDA-native fitting, and the matrix-free `xprod`
route. For large classification tasks, the label-aware PLS-SVD path computes
class-wise cross-products from the factor labels and avoids dense one-hot `Y`
unless the legacy dense route is explicitly needed.

### SIMPLS

`method = "simpls"` is the optimized fastPLS SIMPLS core. Statistically, it
keeps the de Jong (1993) SIMPLS idea: one supervised latent direction is
appended at a time, the score and loading vectors are computed, an orthogonal
loading direction is constructed, and the cross-covariance is deflated before
the next component.

At each refresh, fastPLS extracts fresh candidates from the current deflated
cross-covariance. Depending on the active compiled route, a refresh produces
one direction or a small candidate block. Accepted candidates are always
applied one at a time through the standard SIMPLS orthogonalization and
deflation equations, and the next refresh uses the resulting deflated
operator. CUDA keeps the SIMPLS state on the device. PLS-SVD uses 32
oversampling directions and five power iterations. Accelerated SIMPLS, OPLS,
and kernel-PLS use 32 directions and five iterations for ordinary matrix shapes;
when the explicit cross-covariance would exceed 512 MiB, their automatic
massive-matrix profile requests `oversample = 12` and `power = 1`, while the
executed controls and refresh width are recorded separately. Explicit controls
supplied through `...` take precedence.

The accepted direction is normalized, converted into a SIMPLS component,
orthogonalized, and followed by SIMPLS deflation. A candidate block is consumed
before the next sketch is generated. The
innovation is therefore not a claim of exact de Jong estimator preservation:
it is a
backend-specialized randomized sequential estimator that trades strict
component equivalence for lower computational cost while preserving the
supervised deflation structure.

The retained optimizations reorganize work without altering this sequence. The
cached-deflation update computes \(v^T S\) once and applies
\(S <- S - v(v^T S)\). Coefficients are updated as
\(B_a = B_{a-1} + r_a q_a^T\), and fitted values are updated similarly when
requested. Cross-products can be cached for suitable matrix shapes, prediction
can use compact latent factors, and matrix-free products avoid materializing a
large cross-covariance. PLS-SVD instead uses a one-shot cross-covariance SVD;
OPLS and kernel PLS reuse the same sequential SIMPLS core where applicable.

The complete component path is also a standard feature of reference SIMPLS
software such as `pls::simpls.fit`, which returns coefficient and fitted-value
arrays for components 1 through `ncomp`. The fastPLS contribution is therefore
not the existence of that path, but its compiled incremental execution,
compact prediction, matrix-free products, and backend-specific storage
strategy.
Incremental updates avoid rebuilding \(R_a Q_a^T\) and \(T_a Q_a^T\) from
scratch at requested prefixes; compact prediction stores latent factors instead
of all dense coefficient prefixes; conditional \(X^T X\) caching can benefit
tall matrices; and matrix-free products trade repeated operator applications
for avoiding an explicit \(p\)-by-\(q\) cross-covariance. Because these
trade-offs depend on \(n\), \(p\), \(q\), and the requested component set, they
are memory and execution optimizations rather than a claim of universal
acceleration.

### OPLS

`method = "opls"` follows the orthogonal PLS idea introduced by Trygg and Wold
(2002). It first removes response-orthogonal variation from `X` and then fits
the optimized PLS core to the filtered matrix. The number of orthogonal
components is controlled by `north`. The filtering step estimates orthogonal
scores and loadings, subtracts the corresponding variation from `X`, and stores
the orthogonal filter so the same transformation is applied at prediction time.
With float32 input, stored OPLS matrices use single precision. CUDA performs
OPLS in float32 or float64, while Metal performs it in float32. CUDA keeps
orthogonal filtering, the predictive core, and prediction on device. Metal
uses GPU sample-matrix products and CPU orthogonal filtering, reduced
factorizations, sequential PLS updates, and prediction.
Users select OPLS with `method = "opls"`; the lower-level implementation details
are handled internally.

### Kernel PLS

`method = "kernelpls"` follows the kernel PLS formulation of Rosipal and Trejo
(2001). It constructs a kernel representation before applying the selected
inner PLS model. The supported kernels are `"linear"`, `"rbf"`, and `"poly"`.
For nonlinear kernels, the training kernel matrix is centered and the PLS model
is fitted in kernel space. For the linear kernel, fastPLS avoids unnecessary
kernel materialization and uses the ordinary linear PLS path. For RBF and
polynomial kernels, CUDA constructs, centers, and retains the Gram
representation on device through fitting and prediction. Metal evaluates Gram
and sample-matrix products through Metal and completes reduced factorizations,
sequential updates, and prediction on the CPU. Because this
representation requires quadratic storage in the number of training
observations, an explicit memory guard rejects unsafe workloads before device
allocation.

### Classification heads

For classification, factor responses are treated as PLS-DA targets. The
following heads are classification-specific decoders and are not used for
numeric regression responses. The default
classifier, `classifier = "argmax"`, predicts the class with the largest
predicted dummy-response score. This is the classical PLS-DA decision rule
described in chemometric discrimination work such as Barker and Rayens (2003)
and reviewed by Boulesteix and Strimmer (2007).

The LDA head, `classifier = "lda"`, fits a
regularized linear discriminant model on the PLS latent scores, following the
linear discriminant principle of Fisher (1936). For each class, fastPLS
estimates a score-space mean, an empirical prior, and a common pooled
within-class covariance matrix. If `T` contains the training scores, the pooled
covariance is computed without materializing class-centered score blocks as
`(T^T T - sum_c n_c mu_c mu_c^T) / max(1, n - C)`. The class systems are solved
by Cholesky factorization and triangular solves; fastPLS neither explicitly
inverts the covariance nor uses Gauss-Jordan elimination. Numerical recovery is
deterministic. With `s = trace(Sigma) / q` (or one when this scale is invalid),
the implementation tries `lambda = rho * s` for `rho = 1e-8, 1e-6, 1e-5,
1e-4, 1e-3, 1e-2`, increasing the value only when Cholesky fails. Prediction
uses the standard discriminant
`t^T w_c - 0.5 mu_c^T w_c + log(n_c / n)`. Float32 inputs keep LDA moments,
factorization, coefficients, and prediction buffers in float32. CUDA retains
the score-space covariance, Cholesky factors, triangular-solve workspaces,
coefficients, and prediction buffers on device across component prefixes. The
Metal operation split performs LDA on the CPU using the host-accessible latent
scores returned by the fitting route.


## Method References

The implementation in `fastPLS` is not a line-by-line copy of the papers below;
rather, these papers define the statistical algorithms or numerical building
blocks that the package implements and accelerates.

- Barker, M. and Rayens, W. (2003). Partial least squares for discrimination.
    *Journal of Chemometrics*, 17, 166-173.
- Boulesteix, A.-L. and Strimmer, K. (2007). Partial least squares: a versatile
    tool for the analysis of high-dimensional genomic data. *Briefings in
    Bioinformatics*, 8, 32-44.
- Chong, I.-G. and Jun, C.-H. (2005). Performance of some variable selection
    methods when multicollinearity is present. *Chemometrics and Intelligent
    Laboratory Systems*, 78, 103-112.
- de Jong, S. (1993). SIMPLS: an alternative approach to partial least squares
    regression. *Chemometrics and Intelligent Laboratory Systems*, 18, 251-263.
- Fisher, R. A. (1936). The use of multiple measurements in taxonomic problems.
    *Annals of Eugenics*, 7, 179-188.
- Geladi, P. and Kowalski, B. R. (1986). Partial least-squares regression: a
    tutorial. *Analytica Chimica Acta*, 185, 1-17.
- Halko, N., Martinsson, P.-G., and Tropp, J. A. (2011). Finding structure with
    randomness: probabilistic algorithms for constructing approximate matrix
    decompositions. *SIAM Review*, 53, 217-288.
- Mevik, B.-H. and Wehrens, R. (2007). The pls package: principal component and
    partial least squares regression in R. *Journal of Statistical Software*,
        18,
    1-23.
- Rosipal, R. and Trejo, L. J. (2001). Kernel partial least squares regression
    in reproducing kernel Hilbert space. *Journal of Machine Learning Research*,
    2, 97-123.
- Szymanska, E., Saccenti, E., Smilde, A. K., and Westerhuis, J. A. (2012).
    Double-check: validation of diagnostic statistics for PLS-DA models in
    metabolomics studies. *Metabolomics*, 8, S3-S16.
        doi:10.1007/s11306-011-0330-3.
- Trygg, J. and Wold, S. (2002). Orthogonal projections to latent structures
    (O-PLS). *Journal of Chemometrics*, 16, 119-128.
- Wold, S., Sjostrom, M., and Eriksson, L. (2001). PLS-regression: a basic tool
    of chemometrics. *Chemometrics and Intelligent Laboratory Systems*, 58,
    109-130.

## Session Information

```{r session-info}
sessionInfo()
```
